---
title: "RNA-Seq Data Analysis of Tumor Vs Normal"
author: "Ranjan Jyoti Sarma, PhD"
date: "`r Sys.Date()`"
output: html_document
---

##Rmarkdown
##RNA_Seq Analysis::DEG
```{r Set_Working_Directory, include=FALSE}
setwd("Path/To/Working/Directory")
getwd()
```



#Loading the resquired Libraries
```{r Loading R Libraries}
library("BiocManager")
library("DESeq2")
```



#Reading Meta Data
```{r Loading Meta Data}
m_data = read.table("metadata.txt", header=TRUE)
sample_table = m_data[, c("Sample_ID", "Condition")]
sample_table$Condition = as.factor(sample_table$Condition)
#sample_table$Condition <- relevel(sample_table$Condition, ref = "Control")
sample_table
```


#Reading Count Data
```{r Loading_Count_Matrix}
count_data <- read.table("Raw_counts.txt", header=TRUE, row.names = 1)
count_matrix <- as.matrix(round(counts))
coldata <- data.frame(row.names = colnames(count_matrix), Condition = sample_table$Condition)
```



#Creating DESeq2 object
```{r DESeq dds Object}
dds <- DESeqDataSetFromMatrix(countData = count_matrix, colData = coldata, design = ~Condition)
dds$Condition <- relevel(dds$Condition, ref = "Control") #Important Step to estimate correct Log2FC.
colnames(colData(dds))
```



#Manual Check ppoint#2
```{r optional check 1}
head(dds)
head(assay(dds))
```



#Manual Check point#3
```{r Optional Check 2}
head(assay(dds), 3)
```



##Filtering Low Count so as to avoid low expressed genes 
##keep only those genes with raw counts greater or equal 15 to eleminate very lowly expressed genes
```{r Filtering Step1 }
keep=rowSums(counts(dds)) >= 5
```




#Checking 'keep' object to see genes matching the filtering criteria (counts greater than 5).
```{r Check Keep object}
head(keep, 5)
```



#Keep only the counts greater than 15 counts, Applying Criteria on dds object and transforming to hdds (Filtered) object. Original dds object is intact.
```{r Applying Filtering Criteria, avoiding low expressed genes}
hdds = dds[keep, ]
```




```{r Check before Nomalization}

print(paste("The number of transcripts with low expression (count < 10) removed = ", nrow(assay(dds)) - nrow(assay(hdds))))
print(paste("Number of transcripts remained in 'hdds' object after filtration= ", nrow(assay(hdds))))
```




# Normalization and Measuring DEx
```{r Normalization and Differential Epression Measure}
new_hdds=DESeq(hdds)
```




#Manual check point#5
```{r Check hdds Object}
head(assay(new_hdds))
```



#Write the Normalised Counts of the genes in a file
```{r Saving the normalized counts}
write.csv(counts(new_hdds, normalized = T), file="NormalisedCount.csv", row.names = TRUE)
```




```{r Results of DEx analysis}
#Obtaining the results of differential expression analysis
results = results(new_hdds)
```



#Extracting top 50 Differentially Expressed genes
```{r Top-50 Genes}

#Ordering the results
results_Ordered = results[order(results$pvalue), ]
results_table=data.frame(results_Ordered)

#Total rows in results_table
original_number = nrow(results_table)


#count how many NAs in padj (FDR corrected p-value  for multiple testing)
na_count <- sum(is.na(results_table$padj))
remains = original_number - na_count

 
# Removing rows where the 'padj' column has missing values (NA)
results_table_no_na=results_table[!is.na(results_table$padj),]
head(results_table_no_na)
after_removed_na = nrow(results_table_no_na)


if (remains == after_removed_na) {
  print("Rows with 'padj' with NAs are removed to avoid error in  downstream analysis/Vizualitation.") 
  print(paste("Number of NAs removed:",na_count))
  print(paste("Number of reamaining genes:", remains))
} else {
  print("WARNING:::Rows with 'padj' with NAs are not removed.")
}


#Criteria Given p_value < 0.05 and absolute log2FoldChange >1
sig_genes = subset(results_table_no_na, pvalue < 0.05)

top50 = head(sig_genes, 50)
non_sig = subset(results_table_no_na, pvalue > 0.05)

nrow(non_sig)

num_sig_genes = nrow(sig_genes)
print(paste("Number of Significant DEGs :", num_sig_genes))

```

#Adding HUGO gene symbole instead of Ensembl IDs which come from GTF file during feature/gene count.
```{r Adding Gene Name to the genes}
library(biomaRt)
#Assuming the datasets belong to Human, using gene sets of homo sapiens
ensembl=useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")
all_genes=getBM(attributes=c('ensembl_gene_id', 'external_gene_name'),mart=ensembl)
rownames(all_genes)=all_genes$ensembl_gene_id
top50$gene_name=all_genes[rownames(top50),2]
sig_genes$gene_name=all_genes[rownames(sig_genes),2]
results_table$gene_name=all_genes[rownames(results_table),2]
results_table_no_na$gene_name=all_genes[rownames(results_table_no_na),2]
head(sig_genes)
final_list_50=top50
write.csv(final_list_50, file="Top50_DEGs.csv", row.names = TRUE)
write.table(final_list_50, file="temp_sig_50.txt", sep = ",")
```



#Getting Genes with Absolute log2FC >1
```{r abs Log2FoldChange >1}
abs_lfc_gt_1 = subset(top50,  abs(top50$log2FoldChange) > 1)

num_abs_lfc_gt_1 = nrow(abs_lfc_gt_1)
print(paste("The number of gene of p value < 0.05 and absolute Log2FC> 1:", num_abs_lfc_gt_1))
abs = abs_lfc_gt_1[, c(7,2,3,4,5,6)]
print(abs)
write.csv(abs, file="Absolute_log2FoldChange_gt_1.csv", row.names = TRUE)
```



#Create Directory to save the plots
```{r Plot Directory}
dir.create("Plots")
```



#Visualization 1
```{r Labelled PCA Plot to undestand the variance among the replicates. Quality check before going to downstream interpretation}
library("ggplot2")
library("ggrepel")
vsd=vst(new_hdds, blind=FALSE)
vsd_data=(assay(vsd))
head(vsd)
sample = sample_table$Sample_ID
sample

pca_data = plotPCA(vsd,intgroup="Condition")+geom_text_repel(aes(label=sample))
#Save the plot
pca_data
ggsave(filename = "Plots/PCA_Plot.tiff", plot = pca_data, width = 6, height = 4, units = "in", dpi = 500)
```



#Visualization 2
#bi-clustered HeatMap considering all the 
```{r bi-clustered HeatMap}
head(vsd_data)

#
matched_genes1 <- vsd_data[rownames(vsd_data) %in% rownames(top50), ]
print(matched_genes1)
nrow(matched_genes1)


#matching the only logfc>1 genes
matched_genes2 <- vsd_data[rownames(vsd_data) %in% rownames(abs_lfc_gt_1), ]
print(matched_genes2)
nrow(matched_genes2)



#Replacing sig genes with Log2FC> 1 ensemble IDs with Gene Name
ensemble_ids1 = row.names(top50)
ensemble_ids2 = row.names(abs_lfc_gt_1)

print(ensemble_ids1)
print(ensemble_ids2)


target1 = row.names(matched_genes1)
target2 = row.names(matched_genes2)


ensembl_to_gene_name_map1 = setNames(top50$gene_name, row.names(matched_genes1))
ensembl_to_gene_name_map1

ensembl_to_gene_name_map2 = setNames(abs_lfc_gt_1$gene_name, row.names(matched_genes2))
ensembl_to_gene_name_map2


rownames(matched_genes1) = ensembl_to_gene_name_map1[rownames(matched_genes1)]
matched_genes1
write.csv(matched_genes1, file="selected_top50_for_heat_map.csv", row.names = TRUE)


rownames(matched_genes2) = ensembl_to_gene_name_map2[rownames(matched_genes2)]
matched_genes2
write.csv(matched_genes2, file="selected_logfc_gt_1_for_heat_map.csv", row.names = TRUE)



my_data1 = read.csv("selected_top50_for_heat_map.csv")
my_data2 = read.csv("selected_logfc_gt_1_for_heat_map.csv")
colnames(my_data1)[1] <- "Gene"
colnames(my_data2)[1] <- "Gene"
write.csv(my_data1, file="Significant_Top50_genes_accross_the_samples.csv", row.names=FALSE)
write.csv(my_data2, file="Significant_genes_logfc_gt_1_accross_the_samples.csv", row.names=FALSE)



df <- as.data.frame(colData(vsd))
head(df)



my_data1 <- read.csv("selected_top50_for_heat_map.csv")
colnames(my_data1)[1] <- "Gene"
mat1 <- my_data1[, -1]
gene_names1 <- my_data1[, 1]
head(gene_names1)

library(pheatmap)
heatmap_data1 = pheatmap(mat1, annotation_col = df, labels_row = gene_names1, cluster_cols = TRUE, cuttree_rows = 4, cuttree_cols = 4, annotation_row_names = TRUE,width = 20, height = 5, fontsize = 3)
ggsave(filename ="Plots/heat_map_top50.tiff", plot = heatmap_data1, width = 6, height = 4, units = "in", dpi = 500)




my_data2 <- read.csv("selected_logfc_gt_1_for_heat_map.csv")
colnames(my_data2)[1] <- "Gene"
mat2 <- my_data2[, -1]
gene_names2 <- my_data2[, 1]
head(gene_names2)


heatmap_data2 = pheatmap(mat2, annotation_col = df, labels_row = gene_names2, cluster_cols = TRUE, cuttree_rows = 4, cuttree_cols = 4, annotation_row_names = TRUE,width = 20, height = 5, fontsize = 3)
ggsave(filename ="Plots/heat_map_logfc_gt_1.tiff", plot = heatmap_data2, width = 6, height = 4, units = "in", dpi = 500)


```




#Visualization 3
```{r VolcanoPlot}


#Extracting Log2FC, pvalue and Gene_name

plotting_table=results_table_no_na[ ,c(2,5,7)]
print(plotting_table)
de=plotting_table
print(de)

# Creating a new column 'dge' filled with NA
de$dge <- "NA"
print(de)


de$dge[abs(de$log2FoldChange) < 1] <- "Log2FC<1"
de$dge[de$pvalue < 0.05 & abs(de$log2FoldChange) > 1] <- "p-Value < 0.05 (Log2FC>1)"
de$dge[de$pvalue > 0.05 ] <- "Not Significant"



p1 <- ggplot(data=de, aes(x=log2FoldChange, y=-log10(pvalue), col=dge)) + geom_point()


#From the top 50 DEGs, as the it was ordered and the last p-value was almost '0.00000024', Let's make the y intercept at that point. 
#Let us label only the gene which absolute Log2FC >1 and p-value <0.05.
#The genes above the Y Intercept (Red in Colour) fall within top 50
p2 <- p1 + geom_vline(xintercept=c(-1, 1), col="blue") + geom_hline(yintercept=-log10(0.00000024), col="red") +geom_text_repel(data = subset(de, pvalue < 0.05 & abs(log2FoldChange) > 1), aes(label = gene_name), nudge_x = 0.5, nudge_y = 0.5)
ggsave(filename ="Plots/Volcano_Plot.tiff", plot = p2, width = 6, height = 4, units = "in", dpi = 500)
```
